---
title: "xgxr nlmixr and ggPmx integration"
output:
  html_document:
    self_contained: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
  )

## These options cache the models and the model simulations in R
## To run the actual models on your system, take the save options off.
nlmixrVersion <- as.character(packageVersion("nlmixr"));
options(nlmixr.save=TRUE,
        nlmixr.save.dir=file.path(system.file(package="nlmixr"), nlmixrVersion),
        huxtable.knit_print_df = FALSE);
if (!dir.exists(getOption("nlmixr.save.dir")))
    dir.create(getOption("nlmixr.save.dir"))
```

This shows an example of integrated workflow between `xgxr` `nlmixr`
and `ggPmx`

```{r load}
library(nlmixr)
library(xgxr)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggPMX)

```

# Load the data

```{r}
pkpd_data <- case1_pkpd %>%
  arrange(DOSE) %>%
  select(-IPRED) %>%
  mutate(TRTACT_low2high = factor(TRTACT, levels = unique(TRTACT)),
         TRTACT_high2low = factor(TRTACT, levels = rev(unique(TRTACT))),
         DAY_label = paste("Day", PROFDAY),
         DAY_label = ifelse(DAY_label == "Day 0","Baseline",DAY_label))
 
pk_data <- pkpd_data %>%
  filter(CMT == 2)

pk_data_cycle1 <- pk_data %>%
    filter(CYCLE == 1)
```

# Exploratory analysis using ggplot and xgx helper functions

## Use xgxr for simplified concentration over time, colored by Dose, mean +/- 95% CI

Often in exploring data it is worthwhile to plot by dose by each
nominal time and add the 95% confidence interval. This typical plot
can be cumbersome and lack some nice features that `xgxr` can help
with.  Note the following helper functions:

- `xgx_theme_set()` this sets the theme to black and white color theme
  and other best pratices in `xgxr`.
  
- `xgx_geom_ci()` which creates the Confidence Interval and mean plots
  in a simple interface. 
  
- `xgx_scale_y_log10()` which creates a log-scale that includes the
  minor grids that immediately show the viewer that the plot is a
  semi-log plot without carefully examining the y axis.
  
- `xgx_scale_x_time_units()` which creates an appropriate scale based on
  your times observed and the units you use. It also allows you to
  convert units easily for the right display.
  
- `xgx_annote_status()` which adds a `DRAFT` annotation which is often
  considered best practice when the data or plots are draft.

```{r}
xgx_theme_set() # This uses black and white theme based on xgxr best
                # pratices

# flag for labeling figures as draft
status <- "DRAFT"

time_units_dataset <- "hours"
time_units_plot    <- "days"
trtact_label       <- "Dose"
dose_label         <- "Dose (mg)"
conc_label         <- "Concentration (ng/ml)" 
auc_label          <- "AUCtau (h.(ng/ml))"
concnorm_label     <- "Normalized Concentration (ng/ml)/mg"
sex_label          <- "Sex"
w100_label         <- "WEIGHTB>100"
pd_label           <- "FEV1 (mL)"
cens_label         <- "Censored"


ggplot(data = pk_data_cycle1, aes(x     = NOMTIME,
                                  y     = LIDV,
                                  group = DOSE,
                                  color = TRTACT_high2low)) +
    xgx_geom_ci(conf_level = 0.95) + # Easy CI with xgxr
    xgx_scale_y_log10() + # semi-log plots with semi-log grid minor lines
    xgx_scale_x_time_units(units_dataset = time_units_dataset,
                           units_plot = time_units_plot) +
    # The last line creates an appropriate x scale based on time-units
    # and time unit scale
    labs(y = conc_label, color = trtact_label) +
    xgx_annotate_status(status) #  Adds draft status to plot
```

With this plot you see the mean concentrations confidence intervals
stratified by dose

## Concentration over time, faceted by Dose, mean +/- 95% CI, overlaid on gray spaghetti plots

Not only is it useful to look at the mean concentrations, it is often
useful to look at the mean concentrations and their relationship
between actual individual profiles.  Using `ggplot` coupled with the
`xgxr` helper functions used above, we can easily create these plots
as well:

```{r}
ggplot(data = pk_data_cycle1, aes(x = NOMTIME, y = LIDV)) +
  geom_line(aes(group = ID), color = rgb(0.5, 0.5, 0.5), size = 1, alpha = 0.3) +
  scale_shape_manual(values = c(1, 8)) +
  scale_color_manual(values = c("grey50", "red")) +
  xgx_geom_ci(aes(x = NOMTIME, color = NULL, group = NULL), conf_level = 0.95) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, units_plot = time_units_plot) +
  labs(y = conc_label, color = trtact_label) +
  theme(legend.position = "none") +
  facet_grid(.~TRTACT_low2high) +
  xgx_annotate_status(status)
```

To me it appears the variability seems to be higher with higher doses
and higher with later times.


## Exploring the dose linearity

A common way to explore the dose linearity is to normalize by the
dose.  If the confidence intervals overlap, often this is a dose
linear example.

```{r}
ggplot(data = pk_data_cycle1,
       aes(x = NOMTIME,
           y = LIDV / as.numeric(as.character(DOSE)),
           group = DOSE,
           color = TRTACT_high2low)) +
  xgx_geom_ci(conf_level = 0.95, alpha = 0.5, position = position_dodge(1)) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, units_plot = time_units_plot) +
  labs(y = concnorm_label, color = trtact_label) +
    xgx_annotate_status(status)
```

This example seems to be dose-linear, with the exception of the
censored data. This can be made even more clear by removing the censored data for this plot:

```{r}
ggplot(data = pk_data_cycle1 %>% filter(CENS == 0),
       aes(x = NOMTIME,
           y = LIDV / as.numeric(as.character(DOSE)),
           group = DOSE,
           color = TRTACT_high2low)) +
  xgx_geom_ci(conf_level = 0.95, alpha = 0.5, position = position_dodge(1)) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, units_plot = time_units_plot) +
  labs(y = concnorm_label, color = trtact_label) +
    xgx_annotate_status(status)
```

The lowest dose, with the most censoring, is the one that seems to be
the outlier.  That is likely an artifact of censoring.

Other ways to explore the data include by looking at normalized Cmax
and AUC values (which we will skip in this vignette).

# Exploring Covariates in the dataset

Using the `xgx` helper functions to `ggplot` you can explore the
effect of high baseline weight.  This particular plot is shown below:

```{r}
ggplot(data = pk_data_cycle1, aes(x = NOMTIME,
                                  y = LIDV,
                                  group = WEIGHTB > 100,
                                  color = WEIGHTB > 100)) + 
    xgx_geom_ci(conf_level = 0.95) +
    xgx_scale_y_log10() +
    xgx_scale_x_time_units(units_dataset = time_units_dataset, units_plot = time_units_plot) +
    facet_grid(.~DOSE) +
    labs(y = conc_label, color = w100_label) +
    xgx_annotate_status(status)
```
It seems that the weight effect is not extreme for either dose group

## Summary of exploratory analysis findings

From the exploratory analysis we see:
- The doses seem proportional
- The PK seems to have a 2-compartment model
- Censoring has a large effect on the PK data.

# Fitting the data with nlmixr

First we need to subset to the PK only data and rename `LIDV` to `DV`

```{r}
dat <- case1_pkpd %>%
    rename(DV=LIDV) %>%
    filter(CMT %in% 1:2) %>%
    # Filter (for now) since CENS supprot in nlmixr is in development
    filter(CENS == 0) %>%
    filter(TRTACT != "Placebo")
```

Next create a 2 compartment model:

```{r}

## Use 2 compartment model
cmt2 <- function(){
    ini({
        lka <- log(0.1) # log Ka
        lv <- log(10) # Log Vc
        lcl <- log(4) # Log Cl
        lq <- log(10) # log Q
        lvp <- log(20) # Log Vp

        eta.ka ~ 0.01
        eta.v ~ 0.1
        eta.cl ~ 0.1
        logn.sd = 10
    })
    model({
        ka <- exp(lka + eta.ka)
        cl <- exp(lcl + eta.cl)
        v <- exp(lv + eta.v)
        q <- exp(lq)
        vp <- exp(lvp)
        linCmt() ~ lnorm(logn.sd)
    })
}

## Check parsing
cmt2m <- nlmixr(cmt2)
print(cmt2m)
```

```{r}
## First try log-normal (since the variabilitiy seemed proportional to concentration)
cmt2fit.logn <- nlmixr(cmt2m, dat, "saem", table=tableControl(npde=TRUE,cwres=TRUE))


## Now try proportional
cmt2fit.prop <- cmt2fit.logn %>%
    update(linCmt() ~ prop(prop.sd)) %>%
    nlmixr(est="saem", table=tableControl(npde=TRUE, cwres=TRUE))

## now try add+prop
cmt2fit.add.prop <- cmt2fit.prop %>%
    update(linCmt() ~ prop(prop.sd) + add(add.sd)) %>%
    nlmixr(est="saem", table=tableControl(npde=TRUE, cwres=TRUE))
```

Now that we have run 3 different estimation methods, we can compare the results side-by-side
```{r}
library(huxtable)

as_hux("lognormal"=cmt2fit.logn, "proportional"=cmt2fit.prop, "add+prop"=cmt2fit.add.prop)
```

Note that the additive and proportional model has the additive
component approach zero.  When comparing the objective functions of
log-normal and proportional models, the proportional model has the
lowest objective function value. (Since we modeled log-normal without
data transformation it is appropriate to compare the AIC/Objective
function values)

You may wish to see a visual comparison of the parameters:

```{r}
library(dotwhisker)
## exponentiate=NA causes the parameters to be back-transformed
dwplot(list("lognormal"=cmt2fit.logn, "proportional"=cmt2fit.prop), exponentiate=NA) +
    ## Notice we can use some xgx functions post analysis as well,
    ## since they are ggplot helper functions
    xgx_scale_x_log10() +
    ggtitle("Comparison between lognormal and proportional models",
            "On backtransformed scales")
```

# Model Diagnostics with ggPMX

```{r}
## The controller then can be piped into a specific plot
ctr <- pmx_nlmixr(cmt2fit.logn, conts = c("WEIGHTB"), cats="TRTACT")

ctr %>% pmx_plot_dv_ipred

ctr %>% pmx_plot_dv_pred

ctr %>% pmx_plot_abs_iwres_ipred

## For this display only show 1x1 individual plot for ID 110 for time < 12
ctr %>% pmx_plot_individual(1, filter=ID == 110 & TIME > 0 & TIME < 12, 
                            facets = list(nrow = 1, ncol = 1))

ctr %>% pmx_plot_iwres_dens

ctr %>% pmx_plot_npde_qq

ctr %>% pmx_plot_npde_pred

ctr %>% pmx_plot_npde_time

ctr %>% pmx_plot_eta_qq

## 0-12 hour VPC on semi-log scale
ctr %>% pmx_plot_vpc(filter=TIME > 0 & TIME < 12,
                     scale_y_log10=TRUE, 
                     bin=pmx_vpc_bin("jenks",n=5))

## 0-12 hour VPC on linear scale
ctr %>% pmx_plot_vpc(filter=TIME > 0 & TIME < 12,
                     scale_y_log10=FALSE, 
                     bin=pmx_vpc_bin("jenks",n=5))

ctr %>% pmx_plot_eta_box

ctr %>% pmx_plot_eta_hist

ctr %>% pmx_plot_eta_matrix

## Create a report of all the plots you generated from the controller
if (!file.exists("nlmixr_report.docx")){
    ctr %>% pmx_report("nlmixr_report",".")
}
if (file.exists("nlmixr_report.Rmd"))
    unlink("nlmixr_report.Rmd") # Remove this file so it isn't confused with vignettes
```

This creates 2 reports, both a [pdf](https://github.com/nlmixrdevelopment/nlmixr/raw/master/vignettes/nlmixr_report.pdf) and [word](https://github.com/nlmixrdevelopment/nlmixr/raw/master/vignettes/nlmixr_report.docx) document.


# Simulation of a new scenario with RxODE

By creating events you can simply simulate a new scenario.  Perhaps
your drug development team wants to explore the 100 mg dose 3 times a
day dosing to see what happens with the PK.  You can simply simulate
from the nlmixr model using a [new event
table](https://nlmixrdevelopment.github.io/RxODE/articles/RxODE-event-table.html)
created from RxODE.

In this case we wish to simulate with some variability and see what
happens at steady state:

```{r}
(ev <- et(amt=100, ii=8, ss=1))

ev$add.sampling(seq(0, 8, length.out=100))
print(ev)
```

A nlmixr model already includes information about the parameter
estimates and can simulate without uncertainty in the population
parameters or covariances, like what is done for a VPC. 

If you wish to simulate `100` patients repeated by `100` different
theoretical studies where you simulate from the uncertainty in the
fixed parameter estimates and covariances you can very easily with
nlmixr/RxODE:

```{r}
set.seed(100)
sim1 <- simulate(cmt2fit.logn, events=ev, nSub=100, nStud=100)

print(sim1)
```

You may examine the simulated study information easily, as show in the
`RxODE` printout:

```{r}
head(sim1$thetaMat)
```

You can also see the covariance matricies that are simulated (note
they come from an inverse Wishart distribution):

```{r}
head(sim1$omegaList)
```

```{r}
head(sim1$sigmaList)
```

It is also easy enough to create a plot to see what is going on with
the simulation:

```{r}

p1 <- plot(sim1) ## This returns a ggplot2 object

## you can tweak the plot by the standard ggplot commands
p1 + xlab("Time (hr)") + 
    ylab("Simulated Concentrations of TID steady state")

# And put the same plot on a semi-log plot
p1 + xlab("Time (hr)") + 
    ylab("Simulated Concentrations of TID steady state") +
    xgx_scale_y_log10()
```

For more complex simulations with variability you can also [simulate
dosing windows and sampling
windows](https://nlmixrdevelopment.github.io/RxODE/articles/RxODE-event-table.html#add-doses-and-samples-within-a-sampling-window)
and use any tool you want to summarize it in the way you wish.

